Synthesizing Symmetric Lenses

### 2018-07-09 papers/SymmetricLenses/main.tex
some more changes

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic Diagram for our symmetric lens synthesis algorithm. Regular
    expressions \Regex and \RegexAlt, and a set of examples \Examples, are
    provided as input. \RXSearch searches through stochastic regular expression
    pairs, equivalent to the originals, and proposes them to \GreedySynth.
    \GreedySynth finds a lens, typed between the generated equivalent pairs.
    When \RXSearch determines it has an optimal lens, it returns that lens.}
  \label{fig:high-level-algorithm}
\end{figure}

With simple symmetric lenses and SREs in hand, we develop a new algorithm for
synthesizing likely lenses, based on the previous algorithm for synthesizing
bijective lenses. This algorithm is constructed from \RXSearch and \GreedySynth,
a pair of collaborating search procedures, shown in
Figure~\ref{fig:high-level-algorithm}. Given source and target SREs $S$ and $T$,
the \RXSearch searches for two \emph{related} SREs $S'$ and $T'$ that are
\emph{compatible} (a heuristic that estimates the likelihood our second
algorithm will succeed). $S'$ and $T'$ are related in the sense that they define
the same regular sets as $S$ and $T$ (respectively) and the distribution of
strings is identical. Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression
equivalence, and which we show does not alter the probability distributions of
the languages.

Given the two compatible SREs, \GreedySynth uses a greedy, type-directed
algorithm find a likely lens between them. If algorithm estimates that the lens
it has discovered is more likely than any future lens it can possibly find, the
search terminates. Otherwise, this lens is stored away (if it is the best found
so far) and the search continues with \RXSearch again.

We implemented the \GreedySynth and \RXSearch algorithms and evaluated their
effectiveness of synthesizing synchronization functions on a range of
benchmarks, including data cleaning tasks taken from Flash Fill, information
extraction tasks taken from Augeas, and ad-hoc format synchronization tasks
taken from a variety of locations\footnote{The majority of these format
  synchronization tasks were found from the following index:
  \url{http://fileformats.archiveteam.org/wiki/Electronic_File_Formats}}.

@@

This infrastructure gives us what we need to find likely satisfying lenses from
a collection of satisfying lenses. All such lenses must synchronize data between
the same two SREs $\Regex$ and $\RegexAlt$. For any such lens $\Lens$, we can
calculate the expected number of bits required to recover a string $s$ in
$\LanguageOf{\Regex}$ from a synchronized string $t$ in
$\LanguageOf{\RegexAlt}$. This expected number of bits is just the conditional
entropy of $\Regex$ given $\RegexAlt$ and $\Lens$. To ensure symmetry, the
likelihood we assign to $\Lens$ is the sum of the conditional entropy of
$\Regex$ given $\RegexAlt$ and $\Lens$, \bcp{Am I supposed to know what this
  means?} and the conditional entropy of $\RegexAlt$ given $\Regex$ and
$\Lens$.\bcp{Why is this the right (or a sensible) thing to do, besides the fact
  that it ensures symmetry?}

@@

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work on synthesis~\cite{?,?,?} has shown that we can use
types to dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular expressions,
which come with a rich algebraic structure that makes each regular expression
semantically equal to infinitely many others. To tame that search space and
leverage the benefits of type-directed synthesis, we split our algorithm into
two communicating synthesizers following earlier
work~\cite{quotient-lens-synthesis}. The first synthesizer, \RXSearch, navigates
the space of semantically equivalent regular expressions by iteratively applying
rewrite rules that preserve both the semantics and the probability distributions
of those stochastic regular expressions. \RXSearch ranks pairs of stochastic
regular expressions by the number of rewrite rule applications required to
obtain the pair from the stochastic regular expressions given as input. It
passes the pairs off to the second synthesizer, \GreedySynth, in rank order with
the smallest first.

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the highest
likelihood by performing a type-directed search. In more detail, \GreedySynth
first normalizes the stochastic regular expressions provided by \RXSearch into
\emph{stochastic DNF regular expressions}. These stochastic DNF regular
expressions are, intuitively, stochastic regular expressions with disjunctions
distributed over concatenations, where associativity information has been
removed. Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions and \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names). Finally, \GreedySynth
converts this lens back into a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

As an example of the partnership between \RXSearch and \GreedySynth,
consider the lens shown in Figure~\ref{fig:example_lens}. 
Inferring this lens requires \RXSearch to use a star semiring axiom to
rewrite the regular expression \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into 
\lstinline{"" | (("\n" . emp_salary). ("\n" . emp_salary)$^*$)} 
so that \GreedySynth can find the appropriate lens.

@@

To resolve this tension, \RXSearch uses heuristics to judge whether to
return the current best satisfying lens to the user or to pass the
next set of unrollings to \GreedySynth.
The heuristics favor stopping if the current best satisfying
lens has relatively high likelihood, indicating the lens is very promising.
The heuristics also favor stopping if \RXSearch 
has delievered to \GreedySynth all pairs of stochastic regular
expressions at a given rank and there is a large number of pairs at the next
rank because searching through all such pairs will take a long time.
With this approach, the algorithm can quickly respond with a satisfying
lens with relatively high likelihood.  If users are unhappy with the result, they can refine the
search by supplying additional examples, which serve both to rule out
the previously proposed lens(es) and to reduce the size of the search
space by cutting down on the number of satisfying lenses.

@@

\subsection{Stochastic Regular Expression Equivalences}
$\RXSearch$ needs to be able to find equivalent stochastic regular expressions.
However, existing work does not reason about stochastic regular expression
equivalence. We extend the \emph{star-semiring}~\cite{?} equivalences to
stochastic regular expressions.\bcp{Why these?}

@@

@@

Without unambiguity constraints it is not easy to calculate the entropy. For
example, consider the SRE $\PRegexOr{\lstinline{"a"}}{\lstinline{"a"}}{.5}$.
Using the above computation, this SRE has an entropy of $1$, but the entropy
should be $0$.\bcp{This doesn't show that it's not easy---it just shows that
one particular way of computing it is wrong.} Similar issues occur\bcp{Can we at least hint what they are?} when trying to find the entropy when
$\emptyset$ is a subterm, so we do not define the entropy on $\emptyset$.

In our work, we do not find this to be a problem\bcp{Why might one expect
  it to be a problem, rather than just a technicality?}: we require
unambiguous regular 
expressions as input to our synthesis procedure, and we can easily preprocess
emptysets out of the stochastic regular expressions that are themselves nonempty
by traversing the correct star-semiring equivalences
(\EG, \PRegexOr{\emptyset}{\lstinline{"s"}}{.5} \SSREquiv \lstinline{"s"}).

@@

We now provide the typing derivations and semantics for these lenses. For
simplicity\bcp{Is there any known issue with relaxing this restriction?
  I've often found ti convenient to use id lenses with ambiguous REs...}, we define the typing derivations and semantics only over unambiguous
regular expressions, even though some lenses can be defined over ambiguous
regular expressions (for example, the identity lens doesn't need any ambiguity
constraints).

@@

\subsection{Lens Costs}
Our cost metric is based on the expected amount of information required to
recover a string in one data format from the other. We use the function
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ to bound\bcp{``bound'' or
  ``calculate''?  I.e., is what we calculate correct, or is it only known to
be an upper bound?  If (as it seems) the answer is the latter, then why?}
the expected amount of 
information required to recover a string in $\RegexAlt$ from a string in
$\Regex$, synchronized by $\Lens$. We use the function $\LEntropyOf{\Regex
  \Given \Lens, \RegexAlt}$ to bound the expected amount of information required
to recover a string in $\Regex$ from a string in $\RegexAlt$, synchronized by
$\Lens$.
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\RegexAlt \Given \IdentityLensOf{\RegexAlt}, \RegexAlt}$
    & =
    & [0,0]\\

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Notice, we would be able to calculate
the exact entropy, were it not for \MergeL. If $\MergeLOf{\Lens_1}{\Lens_2}
\OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\Regex$ or $\RegexAlt$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\Regex$ or $\RegexAlt$. In the worst case, the
expected information content to encode this choice is
$-\Log_2(\ProbabilityAlt)*\ProbabilityAlt -\Log_2(1 -
\ProbabilityAlt)*(1-\ProbabilityAlt)$, but the expected information content
could be arbitrarily small.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic. Once that conversion has
taken place, the $\RXSearch$ algorithm executes. This algorithm itself is
defined via five functions: (1) $\PCF{\PQ.Create}$, (2) $\PCF{\PQ.Pop}$, (3)
$\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5) $\PCF{\Continue}$. First,
$\PCF{\PQ.Create}$ creates a priority queue ($pq$) containing a single element:
the initial pair of stochastic regular expressions. Next, $\PCF{SynthSymLens}$
enters a loop that searches for likely lenses. The loop terminates when the
algorithm believes it is unlikely to find a better lens than the best one it has
found so far (a termination condition defined by $\PCF{\Continue}$). Within each
iteration of the loop, it:
\begin{itemize}
\item generates the next class ($S$, $T$) of lenses to
search ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($\PCF{\RXSearch.Best}$).  Each element of this algorithm will be explained in
further depth in the following subsections.

>>>>

\begin{figure}
  \includegraphics[width=.5\textwidth]{high-level-algorithm.pdf}
  \caption{Schematic Diagram for our symmetric lens synthesis algorithm. Regular
    expressions \Regex and \RegexAlt, and a set of examples \Examples, are
    provided as input. \Expand searches through stochastic regular expression
    pairs, equivalent to the originals, and proposes them to \GreedySynth.
    \GreedySynth finds a lens, typed between the generated equivalent pairs.
    When the algorithm determines it has an optimal lens, it returns that lens.}
  \label{fig:high-level-algorithm}
\end{figure}

With simple symmetric lenses and SREs in hand, we develop a new algorithm for
synthesizing likely lenses, based on the previous algorithm for synthesizing
bijective lenses. This algorithm is constructed from \Expand and \GreedySynth,
a pair of collaborating search procedures, shown in
Figure~\ref{fig:high-level-algorithm}. Given source and target SREs $S$ and $T$,
the \Expand searches for two \emph{related} SREs $S'$ and $T'$ that are
\emph{compatible} (a heuristic that estimates the likelihood our second
algorithm will succeed). $S'$ and $T'$ are related in the sense that they define
the same regular sets as $S$ and $T$ (respectively) and the distribution of
strings is identical. Indeed, we guarantee these two properties by implementing
a search procedure that obeys the star semi-ring axioms of regular expression
equivalence, and which we show does not alter the probability distributions of
the languages.

Given the two compatible SREs, \GreedySynth uses a greedy, type-directed
algorithm find a likely lens between them. If algorithm estimates that the lens
it has discovered is more likely than any future lens it can possibly find, the
search terminates. Otherwise, this lens is stored away (if it is the best found
so far) and the search continues with \Expand again.

We implemented the \GreedySynth and \Expand algorithms and evaluated their
effectiveness of synthesizing synchronization functions on a range of
benchmarks, including data cleaning tasks taken from Flash Fill, information
extraction tasks taken from Augeas, and ad-hoc format synchronization tasks
taken from a variety of locations\footnote{The majority of these format
  synchronization tasks were found from the following index:
  \url{http://fileformats.archiveteam.org/wiki/Electronic_File_Formats}}.

@@

This infrastructure gives us what we need to find likely satisfying lenses from
a collection of satisfying lenses. All such lenses must synchronize data between
the same two SREs $\Regex$ and $\RegexAlt$. For any such lens $\Lens$, we can
calculate the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}*\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher cost to lenses where knowing the string present in \Regex
provides little information about what string is present in \RegexAlt. With this
metric, bijections have zero cost -- given a string $s \in \Regex$, we know
exactly what string is present in $\RegexAlt$, as exactly one string is
synchronized to it.

@@

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work on synthesis~\cite{?,?,?} has shown that we can use
types to dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular expressions,
which come with a rich algebraic structure that makes each regular expression
semantically equal to infinitely many others. To tame that search space and
leverage the benefits of type-directed synthesis, we split our algorithm into
two communicating synthesizers following earlier
work~\cite{quotient-lens-synthesis}. The first synthesizer, \Expand, navigates
the space of semantically equivalent regular expressions by applying rewrite
rules that preserve both the semantics and the probability distributions of
those stochastic regular expressions. Our algorithm ranks pairs of stochastic
regular expressions by the number of rewrite rule applications required to
obtain the pair from the stochastic regular expressions given as input. It
passes the pairs off to the second synthesizer, \GreedySynth, in rank order with
the smallest first.

At a high level, \GreedySynth looks for lenses between a given pair of
stochastic regular expressions $\Regex$ and $\RegexAlt$ that have the highest
likelihood by performing a type-directed search. In more detail, \GreedySynth
first normalizes the stochastic regular expressions provided by \Expand into
\emph{stochastic DNF regular expressions}. These stochastic DNF regular
expressions are, intuitively, stochastic regular expressions with disjunctions
distributed over concatenations, where associativity information has been
removed. Then, \GreedySynth uses the syntax of these n-ary DNF regular
expressions to find highly normalized lenses, in a form we call \emph{simple
  symmetric n-ary DNF lenses}. These lenses include neither a composition
operator nor a type equivalence rule. These restrictions mean that each simple
symmetric n-ary DNF lens is typed by a single pair of stochastic n-ary DNF
regular expressions and \GreedySynth's search space for a given pair of regular
expressions is finite. The variations that \GreedySynth considers include
choosing between \IdentityLens{} and \Disconnect lenses and deciding how to
match multiple occurrences of the same kind of data, \EG, the two occurrences of
\lstinline{name} in the management and HR data formats (we want to map first
names to first names and last names to last names). Finally, \GreedySynth
converts this lens back into a simple symmetric lens by converting the n-ary
forms into the binary forms provided in the surface language.

As an example of the partnership between \Expand and \GreedySynth,
consider the lens shown in Figure~\ref{fig:example_lens}. 
Inferring this lens requires \Expand to use a star semiring axiom to
rewrite the regular expression \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into 
\lstinline{"" | (("\n" . emp_salary). ("\n" . emp_salary)$^*$)} 
so that \GreedySynth can find the appropriate lens.

@@

To resolve this tension, our algorithm uses heuristics to judge whether to
return the current best satisfying lens to the user or to pass the
next set of unrollings to \GreedySynth.
The heuristics favor stopping if the current best satisfying
lens has relatively high likelihood, indicating the lens is very promising.
The heuristics also favor stopping if \Expand
has delievered to \GreedySynth all pairs of stochastic regular
expressions at a given rank and there is a large number of pairs at the next
rank because searching through all such pairs will take a long time.
With this approach, the algorithm can quickly respond with a satisfying
lens with relatively high likelihood.  If users are unhappy with the result, they can refine the
search by supplying additional examples, which serve both to rule out
the previously proposed lens(es) and to reduce the size of the search
space by cutting down on the number of satisfying lenses.

@@

\subsection{Stochastic Regular Expression Equivalences}
$\Expand$ needs to be able to find equivalent stochastic regular expressions.
However, existing work does not provide a means to reason about stochastic
regular expression equivalence. We extend the \emph{star-semiring}~\cite{?}
equivalences to stochastic regular expressions.

@@

We prove these star-semiring equivalences sound, as those are the equivalences
we either traverse (with \Expand), or normalize across (as part of
\GreedySynth).

@@

Without unambiguity constraints it is not easy to calculate the entropy. This
method is certainly incorrect without unambiguity constraints: consider the SRE
$\PRegexOr{\lstinline{"a"}}{\lstinline{"a"}}{.5}$. Using the above computation,s
this SRE has an entropy of $1$, but the entropy should be $0$. These
difficulties become particularlly complex when the languages are ambiguous in
subtle ways, the algorithm must be correct when only some of the two languages
overlap. Similar issues occur when trying to find the entropy when $\emptyset$
is a subterm (what should the entropy of $\PRegexStar{\emptyset}{.5}$ be), so we
do not define the entropy on $\emptyset$.

In our work, these conditions are merely a technicality, we require unambiguous
regular expressions as input to our synthesis procedure, and we can easily
preprocess emptysets out of the stochastic regular expressions that are
themselves nonempty by traversing the correct star-semiring equivalences (\EG,
\PRegexOr{\emptyset}{\lstinline{"s"}}{.5} \SSREquiv \lstinline{"s"}).

@@

We now provide the typing derivations and semantics for these lenses. For
simplicity of presentation,\bcp{Is there any known issue with relaxing this
  restriction? I've often found ti convenient to use id lenses with ambiguous
  REs...}\afm{no known issues, I just didn't want to go into specifics about
  ambiguity in an already-full paper}, we define the typing derivations and
semantics only over unambiguous regular expressions, even though some lenses can
be defined over ambiguous regular expressions (for example, the identity lens
doesn't need any ambiguity constraints).

@@

\subsection{Lens Costs}
Our cost metric is based on the expected amount of information required to
recover a string in one data format from the other. We use the function
$\REntropyOf{\RegexAlt \Given \Lens, \Regex}$ to calculate bounds of the
expected amount of information required to recover a string in $\RegexAlt$ from
a string in $\Regex$, synchronized by $\Lens$. We use the function
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ to calculate bounds of the
expected amount of information required to recover a string in $\Regex$ from a
string in $\RegexAlt$, synchronized by $\Lens$.
\begin{center}
  \begin{tabular}{rcl}
    $\REntropyOf{\RegexAlt \Given \IdentityLensOf{\RegexAlt}, \RegexAlt}$
    & =
    & [0,0]\\

@@

    $\REntropyOf{\Regex \Given \InvertOf{\Lens}, \RegexAlt}$
    & =
    & $\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$\\
  \end{tabular}
\end{center}
$\LEntropyOf{\Regex \Given \Lens, \RegexAlt}$ is defined symmetrically. These
functions bound the expected number of bits to recover one data format from a
synchronized string in the other format. Notice, we would be able to exactly
calculate the conditional entropy, were it not for \MergeL. If
$\MergeLOf{\Lens_1}{\Lens_2} \OfType \Regex \Leftrightarrow
\PRegexOr{\RegexAlt_1}{\RegexAlt_2}{\ProbabilityAlt}$, given a string in $s$, we
need to determine if the synchronized string is in $\Regex$ or $\RegexAlt$.
However, this information content is dependent on the how likely the
synchronized string is to be in $\Regex$ or $\RegexAlt$. Furthermore, we
typically calculate the conditional entropy exactly, as merges are relatively
uncommon on practice; only \afm{2} of the lenses sythesized in our benchmarks
include merges.

@@

% begin synthesis
\section{Synthesis}
\label{sec:synthesis}
Algorithm~\ref{alg:synth-sym-lens} presents our synthesis algorithm at a high
level of abstraction. The input regular expressions are first converted into
stochastic regular expressions with \ToStochastic. Once that conversion has
taken place, the we begin searching for the most likely lens. This search
itself is defined via five functions: (1) $\PCF{\PQ.Create}$, (2)
$\PCF{\PQ.Pop}$, (3) $\PCF{Expand}$, (4) $\PCF{\PQ.Push}$, and (5)
$\PCF{\Continue}$. First, $\PCF{\PQ.Create}$ creates a priority queue ($pq$)
containing a single element: the initial pair of stochastic regular expressions.
Next, $\PCF{SynthSymLens}$ enters a loop that searches for likely lenses. The
loop terminates when the algorithm believes it is unlikely to find a better lens
than the best one it has found so far (a termination condition defined by
$\PCF{\Continue}$). Within each iteration of the loop, it:
\begin{itemize}
\item generates the next class ($S$, $T$) of lenses to
search ($\PCF{\PQ.Pop}$),
\item executes $\GreedySynth$ to find a best lens in that class if
one exists ($\LC$),
using the examples $exs$ to filter out potential lenses that do not satisfy
the specification,
\item replaces $best$ with $\LC$, if $\LC$ has lower cost, and
\item adds all of the SREs derived from rewriting $S$ and $T$ ($\Expand(S,T)$)
  to the priority queue with $\PCF{\PQ.Push}$
\end{itemize}
When the loop terminates, the search returns the globally best lens found
($best$).  Each element of this algorithm will be explained in
further depth in the following subsections.
